MLH1 Results Outline

### 2020-02-17 doc/Results.rmd
Finalizing the DMC1 section, hopefully

The same model framwrok was applied to the male specific data.
The male specific analysis was done to assess the variance across strains. The plots below illustrate mouse level means for MLH1 per cell serperated by subspecies.

@@

<!--
interpertation notes,

the number of DMC1 across stages (L and Z stages) are driven by different equations

expectations, predictions if this is the case

@@

DMC1.mean.table

@@

In order to get a good idea on wether variation with mean CO number per cell is associated with the number of percursors upstream in prophase, we quantified DMC1 foci, a marker for DSBs. DMC1 foci were scored from X `r sum( (DMC1.mean.table %>%  filter(stage == "L") )$ncells)` leptotene and `r sum( (DMC1.mean.table %>%  filter(stage == "Z") )$ncells)` zygotene staged spermatocytes of juvenile mice (12-14-18 days). A subset of strains were quantified  _M. m. musculus^PWD^_, _M. m. molossinus^MSM^_, _M. m. musculus^KAZ^_ ,  _M. m. domesticus^WSB^_ , and _M. m. domesticus^G^_. Leptotene and zygotene cells were staged based on SC-AE and centromere morphology.

```{r DMC1.plot, echo=FALSE, warning=FALSE}
#think of better color scheme
#make the jitter col thinner or make the stacked dots.
#remember that box-plots add outlier dots.
#change y labels

@@

DMC1.boxplot <- ggplot(aes(x=strain, y=total.count, color=stage), data=DMC1.DF[DMC1.DF$stage != "P",])+
  geom_boxplot(aes(alpha=.5))+geom_jitter()+ggtitle("DMC1 averages in juvenile mice")+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

@@

<!--
(discription of the nice boxplot figure)

- overall the range is very high (CO homeostasis, technical noise, process is noiser continous repair and new DSBs)

- exclude KAZ?

- expected difference between cell stages for all strains

- pattern of musc strains being higher than the dom strains (subspe pooled t-test)

- MSM is much higher

- the ratios of DMC1 : MLH1 counts!!
-->
```{r plot.stage.ttest, echo=FALSE, warning=FALSE}

@@

the pvalues for the differences between time points are  `r ttest.ZtoL$p.value`,  `r ttest.ZtoL.high$p.value` , and  `r ttest.ZtoL.Low$p.value` for all observations, the high rec group and the low group respectively.

```{r plot.t.test, echo=FALSE}

@@

#high vs low group (L)
ttest.HighLow.L <- t.test(DMC1.DF$total.count[( (DMC1.DF$stage == "L") & ( ( DMC1.DF$strain == "MSM") | (DMC1.DF$strain == "PWD") ) )  ],
               DMC1.DF$total.count[( (DMC1.DF$stage == "L") & ( ( DMC1.DF$strain == "WSB") | (DMC1.DF$strain == "G") | (DMC1.DF$strain == "KAZ") ) )]    )
# p-value =  0.002145

#high vs low group (Z)
ttest.HighLow.Z <- t.test(DMC1.DF$total.count[( (DMC1.DF$stage == "Z") & ( ( DMC1.DF$strain == "MSM") | (DMC1.DF$strain == "PWD") ) )  ],
               DMC1.DF$total.count[( (DMC1.DF$stage == "Z") & ( ( DMC1.DF$strain == "WSB") | (DMC1.DF$strain == "G") | (DMC1.DF$strain == "KAZ") ) )]    )
# p-value = 0.6628

@@

<!--
Subsampling code -- don't report

The number of observations are not quite equal across strains and cell stages, so I ran permutations sampling 25 cells from the high and low recombining groups of cells. Below are the distributions of p values for permuted t-tests from sampling 25 cells from the pooled high and low recombining groups. Permutations were done for both cell stages. The actual p value is in red.

The observed p value does not fall outside of the permuted distribution with assumes equal cell sampling.

```{r DMC1.subsampling, echo=FALSE}

@@

```{r DMC1_MLH1.cor, echo=FALSE, warning = FALSE}
#make category averages from AP_mouse_table.HQ
mean.MLH1.obs <- ddply(.data=AP_mouse_table.HQ,
                 .(category),
                 summarize, 
                 mean.MLH1 = mean(mean_co),
                 var.MLH1 = var(mean_co)
)

##RATIOS
DMC1table.main$MLH1.L_ratio <- (DMC1table.main$mean.count.L / DMC1table.main$mean.MLH1)
DMC1table.main$MLH1.Z_ratio <- (DMC1table.main$mean.count.Z / DMC1table.main$mean.MLH1)

#DMC1.mean.table

as.data.frame(DMC1table.main)

The table above indicates, there is not much variation across the MLH1 : DMC1 ratio (a proxy for the proportion of NCO).

The correlation with MLH1 and leptotene cells is `r MLH1.DMC1.L_cor`. The correlation with MLH1 and zygotene cells is `r MLH1.DMC1.Z_cor`.

`r c(WSB_DMC1.Z_CO_ratio, G_DMC1.Z_CO_ratio, MSM_DMC1.Z_CO_ratio, PWD_DMC1.Z_CO_ratio )` for  _M. m. domesticus^WSB^_ , _M. m. domesticus^G^_, _M. m. molossinus^MSM^_, and  _M. m. musculus^PWD^_ respecitively.

These are the DSB:CO ratios for the L means `r DMC1table.main$MLH1.L_ratio` and Z means `r DMC1table.main$MLH1.Z_ratio`.

```{r DMC1_anova, echo=FALSE, warning=FALSE, eval=FALSE}
#1. ANOVA, is there a difference in DMC1 counts across strains?
mod.L <- lm(total.count ~ strain, data = DMC1.DF[DMC1.DF$stage == "L",])
mod.Z <- lm(total.count ~ strain, data = DMC1.DF[DMC1.DF$stage == "Z",])

l.anova <- anova(mod.L) #0.0002701 ***
z.anova <- anova(mod.Z) #NS

l.anova
z.anova
```

<!--
Since there is evidence for non-equal variance across strains for zygotene cells, so don't rely on lm()s that estimate the effect of strains.

The p-values from the t.tests of the high vs low groups, indicate that the high recombining are significantly higher for the L cells (p value = `r ttest.HighLow.L$p.value`) while the zygotene cells, there is not a significant difference across the high and low groups (value = `r ttest.HighLow.Z$p.value`).
-->

@@

```{r male.show.plotsEXample, echo=FALSE, eval=FALSE}

>>>>

The same model framework was applied to the male specific data.
The male specific analysis was done to assess the variance across strains. The plots below illustrate mouse level means for MLH1 per cell serperated by subspecies.

@@

@@

@@

```{r MAIN.DMC1.plot, echo=FALSE, warning=FALSE}
#think of better color scheme
#make the jitter col thinner or make the stacked dots.
#remember that box-plots add outlier dots.
#change y labels

@@

DMC1.boxplot <- ggplot(aes(x=strain, y=total.count, color=stage), data=DMC1.DF[DMC1.DF$stage != "P",])+
  geom_boxplot(alpha=.5) +geom_jitter()+ggtitle("DMC1 averages in juvenile mice")+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

@@

```{r plot.stage.ttest, echo=FALSE, warning=FALSE}

@@

```{r plot.t.test, echo=FALSE, include=FALSE}

@@

#high vs low group (L)
ttest.HighLow.L.pval <- t.test(DMC1.DF$total.count[( (DMC1.DF$stage == "L") & ( ( DMC1.DF$strain == "MSM") | (DMC1.DF$strain == "PWD") ) )  ],
               DMC1.DF$total.count[( (DMC1.DF$stage == "L") & ( ( DMC1.DF$strain == "WSB") | (DMC1.DF$strain == "G") | (DMC1.DF$strain == "KAZ") ) )]    )$p.value
# p-value =  0.002145

#high vs low group (Z)
ttest.HighLow.Z.pval <- t.test(DMC1.DF$total.count[( (DMC1.DF$stage == "Z") & ( ( DMC1.DF$strain == "MSM") | (DMC1.DF$strain == "PWD") ) )  ],
               DMC1.DF$total.count[( (DMC1.DF$stage == "Z") & ( ( DMC1.DF$strain == "WSB") | (DMC1.DF$strain == "G") | (DMC1.DF$strain == "KAZ") ) )]    )$p.value
# p-value = 0.6628

@@

```{r DMC1_anova, echo=FALSE, warning=FALSE}
#1. ANOVA, is there a difference in DMC1 counts across strains?

#unorder strain
DMC1.DF$strain <- factor(DMC1.DF$strain, levels = c("G","WSB","KAZ","PWD","MSM"), ordered = FALSE )

mod.L <- lm(total.count ~ strain, data = DMC1.DF[DMC1.DF$stage == "L",])
mod.Z <- lm(total.count ~ strain, data = DMC1.DF[DMC1.DF$stage == "Z",])

l.anova.pval <- anova(mod.L)$`Pr(>F)`[1] #0.0002701 ***
z.anova.pval <- anova(mod.Z)$`Pr(>F)`[1] #NS
```

```{r DMC1_MLH1.MAIN_TABLE, echo=FALSE, warning = FALSE}
#make category averages from AP_mouse_table.HQ
mean.MLH1.obs <- ddply(.data=AP_mouse_table.HQ,
                 .(category),
                 summarize, 
                 mean.MLH1 = mean(mean_co),
                 var.MLH1 = var(mean_co)
)

##RATIOS
DMC1table.main$MLH1.L_ratio <- DMC1table.main$mean.MLH1/ DMC1table.main$mean.count.L 
DMC1table.main$MLH1.Z_ratio <- DMC1table.main$mean.MLH1 / DMC1table.main$mean.count.Z

#colnames(DMC1table.main) <- c("category", "mean.MLH1", "", "", "")

#DMC1.mean.table
DMC1.mean.table.99 <- subset(DMC1table.main, select= -c(var.MLH1,  mean.count.x,  mean.count.y, stage.x, stage.y))

colnames(DMC1.mean.table.99) <- c("category","mean.MLH1","ncells.L","mean.L",
                  "ncells.Z", "mean.Z", "MLH1.L_ratio", "MLH1.Z_ratio")

DMC1.mean.table.99

DMC1table.main$category <- factor(DMC1table.main$category, levels = c("WSB male","G male","KAZ male",               "PWD male","MSM male"), ordered = TRUE )

DMC1table.main <- DMC1table.main[order(DMC1table.main$category),]
as.data.frame(DMC1table.main)

DMC1.Lratio_t.test <- t.test(  (DMC1table.main %>% filter(  ( category == "G male" | category == "WSB male" | category == "KAZ male" ) ) )$MLH1.L_ratio, 
                (DMC1table.main %>% filter(  ( category == "PWD male" | category == "MSM male") ) )$MLH1.L_ratio        )

DMC1.Zratio_t.test <- t.test(  (DMC1table.main %>% filter(  ( category == "G male" | category == "WSB male" | category == "KAZ male" ) ) )$MLH1.Z_ratio, 
                (DMC1table.main %>% filter(  ( category == "PWD male" | category == "MSM male") ) )$MLH1.Z_ratio        )

In an attempt to locatize the above male specific crossoever number evolution within the meiotic pathway we quantified a marker for DSBs, DMC1, in early prophase meiocytes.

DMC1 foci were scored from a total of `r sum( (DMC1.mean.table %>%  filter(stage == "L") )$ncells)` leptotene and `r sum( (DMC1.mean.table %>%  filter(stage == "Z") )$ncells)` zygotene staged spermatocytes of (X for each strain) juvenile mice (12 to 18 days).

A subset of strains were choosen to test for differences between the high and low recombination groups; _musculus^PWD^_, _molossinus^MSM^_, _ musculus^KAZ^_ ,  _domesticus^WSB^_ , and _domesticus^G^_. One mouse was sampled for each strain.

As can be seen in the figure, the high recombining groups have significantly more DMC1 for for leptotene (t.test; `r ttest.HighLow.L.pval` , one-way-anova; `r round(l.anova.pval, 5)`), but not the zygotene cells (t.test; `r ttest.HighLow.Z.pval`, one-way-anova; `r z.anova.pval`).

The ratios for MLH1:DMC1^Leptotene^ and MLH1:DMC1^zygotene^ are shown in Table X, and can be used as a proxy for the CO:NCO ratio (instead CO:(CO+NCO) ratio). Notably there is not much variation and do not show a significant difference between the high and the low recombining groups (t.test; p = `r DMC1.Lratio_t.test$p.value` and `r DMC1.Zratio_t.test$p.value` for leptotene and zygotene based ratios respectively). These results indicate that the best predictor for CO number may be established before DSB formation (cite Baier et al 2014).

<!--
The correlation with MLH1 and leptotene cells is `r MLH1.DMC1.L_cor`. The correlation with MLH1 and zygotene cells is `r MLH1.DMC1.Z_cor`. `r c(WSB_DMC1.Z_CO_ratio, G_DMC1.Z_CO_ratio, MSM_DMC1.Z_CO_ratio, PWD_DMC1.Z_CO_ratio )` for  _M. m. domesticus^WSB^_ , _M. m. domesticus^G^_, _M. m. molossinus^MSM^_, and  _M. m. musculus^PWD^_ respecitively. 
-->

```{r DMC1.subsampling, echo=FALSE, eval = FALSE, include = FALSE}

#Subsampling code -- don't report

#The number of observations are not quite equal across strains and cell stages, so I ran permutations sampling 25 cells from the high and low recombining groups of cells. Below are the distributions of p values for permuted t-tests from sampling 25 cells from the pooled high and low recombining groups. Permutations were done for both cell stages. The actual p value is in red.

#The observed p value does not fall outside of the permuted distribution with assumes equal cell sampling.

@@

<!--
interpertation notes, 
the number of DMC1 across stages (L and Z stages) are driven by different equations

expectations, predictions if this is the case -- more within mouse/group variation for DMC1 number in Z compared to L?
OR

variance in L stage for DMC1 per cell would be proportional to the variation in the 'packing ratio' across pachtyene cells
-Baier et al 2014 supports this model (Chmr axis / DNA loop model)
-->

@@

```{r male.show.plotsEXample, echo=FALSE, eval=FALSE}
male_mouse.av <- subset(AP_mouse_table_w.Ages, select= -c(sex.y,sex.x,
                            subsp.y,subsp.x,
                    category.y,category.x, strain.x,
                    strain.y, mouse.1  )  )
